---
title: 'Final Proposal'
author: "Roba Abbajabal, Conor O'Shea, Aadhi Subbiah, Daniel Nass"
date: "11/11/2023"
output: html_document
---

# Research Title: An Analysis on Transportation Safety
##### Team Members: Roba Abbajabal, Conor O'Shea, Aadhi Subbiah, Daniel Nass

#### Description of Data: 
This data set contains information from 2010-2021 regarding fatal vehicular accidents that have occured on highways within the United States by state/territory.
The data can be acessed at the following link: https://www.bts.gov/browse-statistical-products-and-data/state-transportation-statistics/transportation-safety
The varibles included in this dataset are as follows: 
- State: the state that the information comes from
- Fatality Type: A factor of 14 different types of fatalities (such as highway fatalities, SUV fatalities, etc.)
- Year: The year, from 2010 to 2021, of the number of a specific type of fatality in a state occured.

#### Investigation: 
What states, if any, have stark contrasts in vehicular fatalities? Do these states have any laws or regulations that would cause this?
Have counts of certain types of vehicular fatalities increased or decreased over time?
Do any types of vehicular fatalities correlate with another?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# csv being used
library(tidyverse)
library(dplyr)

hw <- read.csv("hwFatal.csv")
head(hw)

# information about data set
summary(hw)
```

```{r}
# Cleaning data
## Fixing empty column name for Fatality Type
hw <-  read.csv("hwFatal.csv", col.names = c("State","FatalityType","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019","2020","2021"))

## Creating a seocnd dataset where there the whole country collectively is not included (just states and territories individually)
states <- hw %>% filter(State != "United States")
head(states)

# information about data set
summary(states)
```