---
title: 'Final Proposal'
author: "Roba Abbajabal, Conor O'Shea, Aadhi Subbiah, Daniel Nass"
date: "11/11/2023"
output: html_document
---

# Research Title: An Analysis on Transportation Safety
##### Team Members: Roba Abbajabal, Conor O'Shea, Aadhi Subbiah, Daniel Nass

#### Description of Data: 
This data set contains information from 2010-2021 regarding fatal vehicular accidents that have occured on highways within the United States by state/territory.
The data can be acessed at the following link: https://www.bts.gov/browse-statistical-products-and-data/state-transportation-statistics/transportation-safety
The varibles included in this dataset are as follows: 
- State: the state that the information comes from
- Fatality Type: A factor of 14 different types of fatalities (such as highway fatalities, SUV fatalities, etc.)
- Year: The year, from 2010 to 2021, of the number of a specific type of fatality in a state occured.

#### Investigation: 
What states, if any, have stark contrasts in vehicular fatalities? Do these states have any laws or regulations that would cause this?
Have counts of certain types of vehicular fatalities increased or decreased over time?
Do any types of vehicular fatalities correlate with another?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

```{r}
# csv being used
hw <- read.csv("hwFatal.csv")
head(hw)

# information about data set
summary(hw)
```

```{r}
# Cleaning data

## Fixing empty column name for Fatality Type
hw <-  read.csv("hwFatal.csv", col.names = c("State","FatalityType","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019","2020","2021"))

## Creating a seocnd dataset where there the whole country collectively is not included (just states and territories individually)
states <- hw %>% filter(State != "United States")
states$State <- tolower(states$State) # making joining with the map easier later on
head(states)

# information about data set
summary(states)
```

```{r}
# Aadhi

# Initializing stuff
state_map <- map_data("state")

colnames(state_map)[5] <- "State"   # making join easier

states_year <- states %>%        # Organizing by year and the count a numeric to plot
  pivot_longer(cols = X2010:X2021, 
               names_to = "year", 
               values_to = "count") %>% 
  mutate(`count` = as.numeric(`count`))

temp_states <- left_join(state_map, states_year, by = "State") # making df to plot




# Pedestrian Fatalities by Year
states_year %>% 
  filter(FatalityType == "Pedestrian Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col() + 
  ggtitle("Pedestrian Fatalities by Year")

# Pedestrian Fatalities by State
temp_states %>% 
  filter(FatalityType == "Pedestrian Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Pedestrian Fatalities by State")






# Other Nonoccupant Fatalities by Year
states_year %>% 
  filter(FatalityType == "Other Nonoccupant Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col() + 
  ggtitle("Other Nonoccupant Fatalities by Year")

# Other Nonoccupant Fatalities by State
temp_states %>% 
  filter(FatalityType == "Other Nonoccupant Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Other Nonoccupant Fatalities by State")






# Other Vehicle Fatalities by Year
states_year %>% 
  filter(FatalityType == "Other Vehicle Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col() + 
  ggtitle("Other Vehicle Fatalities by Year")

# Other Vehicle Fatalities by State
temp_states %>% 
  filter(FatalityType == "Other Vehicle Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Other Vehicle Fatalities by State")
```
```{r}
temp_states <- left_join(state_map, states_year, by = "State")

temp_states %>% 
  filter(FatalityType == "Pedestrian Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities))
  

```

``` {r, Warnings=FALSE}
# Daniel

# Bus Fatalities by year per state
states_year %>% 
  filter(FatalityType == "Bus Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Bus Fatalities by Year")

temp_states %>%
  filter(FatalityType == "Bus Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = count)) + 
  ggtitle("Bus Fatalities by State")


# Large Truck Fatalities
states_year %>% 
  filter(FatalityType == "Large Truck Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Large Truck Fatalities by Year")

temp_states %>%
  filter(FatalityType == "Large Truck Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = count)) + 
  ggtitle("Large Truck Fatalities by State")


# Pedalcyclist Fatalities
states_year %>% 
  filter(FatalityType == "Pedalcyclist Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Pedalcyclist Fatalities by Year")

temp_states %>%
  filter(FatalityType == "Pedalcyclist Fatalities") %>% 
  mutate(total_fatalities = sum(count), .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = count)) + 
  ggtitle("Pedalcyclist Fatalities by State")

```

