---
title: 'Final Report'
author: "Roba Abbajabal, Conor O'Shea, Aadhi Subbiah, Daniel Nass"
date: "11/11/2023"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

### Introduction

-   For our research, we wished to analyze transportation safety in the
    US by state from 2010-2021.
-   This involved observing fatalities by state and year, and then
    connecting that information with outside sources.
-   This research particularly interests us, because we wish to find
    reasons of what may cause safer roads.

### Research Title: An Analysis on Transportation Safety in the US

### Team Members: Roba Abbajabal, Conor O'Shea, Aadhi Subbiah, Daniel Nass

#### Description of Data:

This data set contains information from 2010-2021 regarding fatal
vehicular accidents that have occurred on highways within the United
States by state/territory. The data can be accessed at the following
link:
<https://www.bts.gov/browse-statistical-products-and-data/state-transportation-statistics/transportation-safety>
The variables included in this dataset are as follows:

-   State: the state that the information comes from

-   Fatality Type: A factor of 14 different types of fatalities (such as
    highway fatalities, SUV fatalities, etc.)

    -   Highway Fatalities

    -   Passenger Car Fatalities

    -   SUV Fatalities

    -   Pickup Truck Fatalities

    -   Van Fatalities

    -   Other Light Truck Fatalities

    -   Motorcycle Fatalities

    -   Bus Fatalities - Large Truck Fatalities

    -   Pedalcyclist Fatalities

    -   Pedestrian Fatalities

    -   Other Nonoccupant Fatalities

    -   Other Vehicle Fatalities

    -   Unknown Person Type Fatalities

-   Year: The year, from 2010 to 2021, of the number of a specific type
    of fatality in a state occured.

-   TotalFatalities: the sum of all fatalities for each FatalityType and
    by year for overall and by each state.

#### Supplementary Datasets Used:

**State Population Totals** (2010-2020)
[state_populations_2010-2021.csv]: This dataset from Bureau of
Transportation Safety contains information from 2010-2021 regarding
fatal vehicular accidents that have occurred on highways within the
United States by state/territory. The data can be accessed at the
following link:
<https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-state-total.html>
This dataset will be used for reporting the number of fatalities by
state per capita. The variables included in this dataset are as follows:

-   Name: The state that information comes from.

-   POPESTIMATE([2010-2021]): The population estimates for year
    ([2010-2021]).

**State Highway Travel** (2010-2021)
[vehicles_traveleld_licensed_drivers_2010_2021.csv]: The Highway
Statistics Series consists of annual reports containing analyzed
statistical information on motor fuel, motor vehicle registrations,
driver licenses, highway user taxation, highway mileage, travel, and
highway finance. These information are presented in tables as well as
selected charts. It has been published annually since 1945. This data
can be accessed using the following link:
<https://www.bts.gov/browse-statistical-products-and-data/state-transportation-statistics/state-highway-travel>
This dataset will be used for reporting the number of miles driven each
year and the number of licensed drivers per state The variables that are
being used are as follows:

-   State: the name of the state the information is from

-   Measures: A factor of 2 different types of data

-   Licensed drivers - Highway vehicle-miles traveled (millions)

-   Year of Year: year the information is from (2010-2021)

-   Values: value of measure factor

**State Financials (2019)** [ASFIN_FY2019.csv]: This dataset from
Census.gov contains information in year 2019 regarding state revenue and
expenditure from different sources. We used this data to show what
correlation traffic fatalities may have with state financial data in the
year 2019. The year 2019 was chosen because it was the year before
COVID-19 arrived to the United States, which we later saw had it's own
impact on traffic fatalities. The variables in the dataset (after using
pivot_longer) are as follows:

-   State: The name of the state the information is collected form.

-   Amount (Per 1000): The amount per \$1000 dollars of state revenue or
    expenditure per category.

-   RevenueType: The type of revenue or expenditure being reported.

#### Investigations:

-   What is the leading cause of highway fatalities overall/by state?
-   What states, if any, have stark contrasts in vehicular fatalities?
-   Do these states have any laws or regulations that would cause this?
-   Do any types of vehicular fatalities correlate with another?
-   Have counts of certain types of vehicular fatalities increased or
    decreased over time?
-   How did the COVID-19 pandemic during 2021 affect traffic fatalities?

#### Main:

The data does not account directly for population density. The data is
raw numbers, and in order to get a better insight into what the data
means we have to look at it by state.

### Data Setup

```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

#### Highway Fatalities dataset being used

```{r, echo=FALSE, warning=FALSE}
# Datasets being used
hw <- read.csv("hw_fatal.csv")
head(hw)
```

#### Summary of the Highway Fatalities dataset

```{r, echo=FALSE, warning=FALSE}
# information about data set
summary(hw)
```

#### State Populations dataset form 2010-2021 being used

```{r, echo=FALSE, warning=FALSE}
statePop <- read.csv("state_populations_2010-2021.csv", col.names = c("State", "Y2010","Y2011","Y2012","Y2013","Y2014","Y2015","Y2016","Y2017","Y2018","Y2019","Y2020","Y2021"))
head(statePop)
```

#### Summary of the State Populations dataset

```{r, echo=FALSE, warning=FALSE}
# information about dataset
summary(statePop)
```

#### State Highway Travel dataset: 2010-2021 being used

```{r, echo=FALSE, warning=FALSE}

hw_travel <- read.csv("vehicles_traveled_licensed_drivers_2010_2021.csv", col.names = c("State", "Type", "Year", "Count"))
head(hw_travel)
```

#### Summary of the State Highway Travel dataset

```{r, echo=FALSE, warning=FALSE}
# information about dataset
summary(hw_travel)
```

### Data Cleanup:

```{r, echo=FALSE, warning=FALSE}
# Cleaning data

## Fixing empty column name for Fatality Type, and making year column names start with a character
hw <-  read.csv("hw_fatal.csv", col.names = c("State","FatalityType","Y2010","Y2011","Y2012","Y2013","Y2014","Y2015","Y2016","Y2017","Y2018","Y2019","Y2020","Y2021","TotalFatalities"))

## Creating a seocnd dataset where there the whole country collectively is not included (just states and territories individually)
states <- hw %>% filter(State != "United States")
states$State <- tolower(states$State) # making joining with the map easier later on
```

#### Creating a second dataset where there the whole country collectively is not included (just states and territories individually)

```{r, echo=FALSE, warning=FALSE}
head(states)

## Parse the fatality by year as a number, and not a character.
states <- states %>%
    mutate (
        Y2010 = parse_number(Y2010),
        Y2011 = parse_number(Y2011),
        Y2012 = parse_number(Y2012),
        Y2013 = parse_number(Y2013),
        Y2014 = parse_number(Y2014),
        Y2015 = parse_number(Y2015),
        Y2016 = parse_number(Y2016),
        Y2017 = parse_number(Y2017),
        Y2018 = parse_number(Y2018),
        Y2019 = parse_number(Y2019),
        Y2020 = parse_number(Y2020),
        Y2021 = parse_number(Y2021),
        TotalFatalities = parse_number(TotalFatalities)
    )
```

#### Summary of newly created states dataset

```{r, echo=FALSE, warning=FALSE}
# information about data set
summary(states)
```

### Data Exploration

#### Plan of Data Exploration

Break down each fatality type and plot the data by state, observe
similarities and differences between the states for each fatality type.

#### Construct dataframe of fatality by year

```{r, echo=FALSE, warning=FALSE}
# Aadhi

# Initializing maps
state_map <- map_data("state")

colnames(state_map)[5] <- "State"   # making join easier

states_year <- states %>%        # Organizing by year and the count a numeric to plot
  pivot_longer(cols = Y2010:Y2021, 
               names_to = "year", 
               values_to = "count") %>% 
  mutate(`count` = as.numeric(`count`))

states_pops <- statePop %>%        # Organizing by year and the count a numeric to plot
  pivot_longer(cols = Y2010:Y2021, 
               names_to = "year", 
               values_to = "population") %>% 
  mutate(`population` = as.numeric(`population`))


states_pops$State <- tolower(states_pops$State)

temp_states_old <- left_join(state_map, states_year, by = "State") # making df to plot
temp_states <- left_join(temp_states_old, states_pops, by = c("State", "year"))

# Roba
## Construct map of fatalality by year

state_map_year <- states %>% 
  pivot_longer(cols = Y2010:Y2021, 
               names_to = "year", 
               values_to = "count") %>% 
  mutate(`count` = as.numeric(`count`))

```

#### Construct dataframe of fatality by state

```{r, echo=FALSE, warning=FALSE}

## Construct map of fatalality by state (Roba)

states$region <- tolower(states$State)
state_map2 <- map_data("state")
state_map2 <- states %>% left_join(state_map2, by="region")
head(state_map2)
```

#### Construct dataframe of Highway Fatalities by Year and State

```{r, echo=FALSE, warning=FALSE}
# Overall highway fatalities by year and state
temp_states %>%
  filter(FatalityType == "Highway Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Highway Fatalities by State per Capita")

state_map2 %>% filter(FatalityType == "Highway Fatalities") %>% 
    mutate(total_fatalities = TotalFatalities, .by = State) %>%   
    ggplot(aes(x = long, y = lat)) + 
    geom_polygon(aes(group=group, fill = total_fatalities)) +
    scale_fill_continuous(name = "Death Count (2010 - 2021)") +
    ggtitle("Highway Fatalities by State (2010-2021)")

states_year %>% 
  filter(FatalityType == "Highway Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Highway Fatalities by Year by State") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

state_map_year %>% 
  filter(FatalityType == "Highway Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("Highway Fatality Totals by Year")

```

#### Comparing highway fatalities to miles traveled in each state/overall from 2010-2021

```{r, echo=FALSE, warning=FALSE}

# Show the number of combined highway vehicle miles traveled each year (2010-2021)
hw_travel %>% 
  filter(Type == "Highway vehicle-miles traveled (millions)") %>% 
  ggplot(aes(x = Year, y = Count, fill = State)) + 
  geom_col() + 
  ggtitle("Highway vehicle-miles traveled (millions) by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# Show the number of total licensed drivers in the US by year
hw_travel %>% 
  filter(Type == "Licensed drivers") %>% 
  ggplot(aes(x = Year, y = Count, fill = State)) + 
  geom_col() + 
  ggtitle("Licensed drivers by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

```{r, echo=FALSE, warning=FALSE}
# Aadhi

# Pedestrian Fatalities by Year
states_year %>% 
  filter(FatalityType == "Pedestrian Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col() + 
  ggtitle("Pedestrian Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Pedestrian Fatalities by State
temp_states %>% 
  filter(FatalityType == "Pedestrian Fatalities") %>% 
   mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Pedestrian Fatalities by State")

# Pedestrian Fatalities by State per capita
temp_states %>% 
  filter(FatalityType == "Pedestrian Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Pedestrian Fatalities by State")



# Other Nonoccupant Fatalities by Year
states_year %>% 
  filter(FatalityType == "Other Nonoccupant Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col() + 
  ggtitle("Other Nonoccupant Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Other Nonoccupant Fatalities by State

temp_states %>% 
  filter(FatalityType == "Other Nonoccupant Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Other Nonoccupant Fatalities by State")

temp_states %>% 
  filter(FatalityType == "Other Nonoccupant Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Other Nonoccupant Fatalities by State")

# Other Vehicle Fatalities by Year
states_year %>% 
  filter(FatalityType == "Other Vehicle Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col() + 
  ggtitle("Other Vehicle Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Other Vehicle Fatalities by State
temp_states %>% 
  filter(FatalityType == "Other Vehicle Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Other Vehicle Fatalities by State")

temp_states %>% 
  filter(FatalityType == "Other Vehicle Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Other Vehicle Fatalities by State")
```

```{r, echo=FALSE, warning=FALSE}
# Conor

# Pickup Truck Fatalities by Year (Conor)

states_year %>% 
  filter(FatalityType == "Pickup Truck Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("Pickup Truck Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
# Pickup Truck Fatalities by State (Conor)

temp_states %>% 
  filter(FatalityType == "Pickup Truck Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Pickup Truck Fatalities by State")

temp_states %>% 
  filter(FatalityType == "Pickup Truck Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Pickup Truck Fatalities by State")

# Van Fatalities by Year (Conor)

states_year %>% 
  filter(FatalityType == "Van Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("Van Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Van Fatalities by State (Conor)

temp_states %>% 
  filter(FatalityType == "Van Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Van Fatalities by State")

temp_states %>% 
  filter(FatalityType == "Van Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Van Fatalities by State")

# Other Light Truck Fatalities by Year (Conor)

states_year %>% 
  filter(FatalityType == "Other Light Truck Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("Other Light Truck Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Other Light Truck Fatalities by State (Conor)

temp_states %>% 
  filter(FatalityType == "Other Light Truck Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Other Light Truck Fatalities by State")

temp_states %>% 
  filter(FatalityType == "Other Light Truck Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Other Light Truck Fatalities by State")

# Motorcycle Fatalities by Year (Conor)

states_year %>% 
  filter(FatalityType == "Motorcycle Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("Motorcycle Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Motorcycle Fatalities by State (Conor)

temp_states %>% 
  filter(FatalityType == "Motorcycle Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Motorcycle Fatalities by State")

temp_states %>% 
  filter(FatalityType == "Motorcycle Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Motorcycle Fatalities by State")


```

```{r, echo=FALSE, warning=FALSE}
# Daniel

# Bus Fatalities by year per state
states_year %>% 
  filter(FatalityType == "Bus Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Bus Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

temp_states %>%
  filter(FatalityType == "Bus Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Bus Fatalities by State")

temp_states %>%
  filter(FatalityType == "Bus Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Bus Fatalities by State")


# Large Truck Fatalities by year per state
states_year %>% 
  filter(FatalityType == "Large Truck Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Large Truck Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

temp_states %>%
  filter(FatalityType == "Large Truck Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Large Truck Fatalities by State")

temp_states %>%
  filter(FatalityType == "Large Truck Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Large Truck Fatalities by State")

# Pedalcyclist Fatalities by year per state
states_year %>% 
  filter(FatalityType == "Pedalcyclist Fatalities") %>% 
  ggplot(aes(x = year, y = count, fill = State)) + 
  geom_col(space=1) + 
  ggtitle("Pedalcyclist Fatalities by Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

temp_states %>%
  filter(FatalityType == "Pedalcyclist Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Pedalcyclist Fatalities by State per Capita")

temp_states %>%
  filter(FatalityType == "Pedalcyclist Fatalities") %>% 
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities)) + 
  ggtitle("Pedalcyclist Fatalities by State")

```

```{r, echo=FALSE, warning=FALSE}
# Roba
  
# Passenger Car Fatalities by State per Capita
temp_states %>%
  filter(FatalityType == "Passenger Car Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("Passenger Car Fatalities by State per Capita")

Passenger Car Fatalities by State
state_map2 %>% filter(FatalityType == "Passenger Car Fatalities") %>%
    mutate(total_fatalities = TotalFatalities, .by = State) %>%    
    ggplot(aes(x = long, y = lat)) + 
    geom_polygon(aes(group=group, fill = total_fatalities)) +
    scale_fill_continuous(name = "Death Count (2010 - 2021)") +
    ggtitle("Passenger Car Fatalities by State")

# SUV Fatalities by State per Capita
temp_states %>%
  filter(FatalityType == "SUV Fatalities") %>% 
  mutate(total_pop = sum(population), .by = State) %>%
  mutate(total_fatalities = TotalFatalities, .by = State) %>% 
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon(aes(group = group, fill = total_fatalities / total_pop)) + 
  ggtitle("SUV Fatalities by State per Capita")

# SUV Fatalities by State
state_map2 %>% filter(FatalityType == "SUV Fatalities") %>% 
    mutate(total_fatalities = TotalFatalities, .by = State) %>%    
    ggplot(aes(x = long, y = lat)) + 
    geom_polygon(aes(group=group, fill = total_fatalities)) +
    scale_fill_continuous(name = "Death Count (2010 - 2021)") +
    ggtitle("SUV Fatalities by State")

# Passenger Car Fatalities by Year
state_map_year %>% 
  filter(FatalityType == "Passenger Car Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("Passenger Car Fatalities by Year")

# SUV Fatalities by Year
state_map_year %>% 
  filter(FatalityType == "SUV Fatalities") %>% 
  ggplot(aes(x = year, y = count)) + 
  geom_col() + 
  ggtitle("SUV Fatalities by Year")

# Other Nonoccupant Fatalities by State
state_map2 %>% filter(FatalityType == "Other Nonoccupant Fatalities") %>% 
    mutate(total_fatalities = TotalFatalities, .by = State) %>%    
    ggplot(aes(x = long, y = lat)) + 
    geom_polygon(aes(group=group, fill = total_fatalities)) +
    scale_fill_continuous(name = "Death Count (2010 - 2021)") +
    ggtitle("Other Nonoccupant Fatalities by State")
```

### Investigating Correlations to State Revenue & Expenditure

#### Reading and Formatting the Financial Data

```{r}
financials2019_old1 <- read.csv("ASFIN_FY2019.csv")
head(financials2019_old1)

financials2019_old2 <- financials2019_old1 %>%        # Organizing by year and the count a numeric to plot
  pivot_longer(cols = !RevenueType, 
               names_to = "State", 
               values_to = "Amount (Per 1000)") %>% 
  mutate(`Amount (Per 1000)` = parse_number(`Amount (Per 1000)`))
financials2019 <- financials2019_old2
financials2019$State <- tolower(financials2019_old2$State)
financials2019$State <- gsub(".", " ", financials2019$State, fixed=TRUE)
head(financials2019)
```

#### Joining Financial Data with Traffic Fatality Data

```{r}
fat_states_pops_2019 <- temp_states %>% filter(year == "Y2019")
head(fat_states_pops_2019)

fat_states_pops_fin_2019 <- left_join(fat_states_pops_2019, financials2019, by = "State")
head(fat_states_pops_fin_2019)
```

#### Investigating Correlations of Traffic Fatalities with Total State Revenue

```{r}
highwayFats_states_pops_revenue_2019 <-
    fat_states_pops_fin_2019 %>% 
    filter(RevenueType == "Total revenue", FatalityType == "Highway Fatalities") %>% 
    select(group, long, lat, State, FatalityType, TotalFatalities, `Amount (Per 1000)`, population) %>%
    distinct()

highwayFats_states_pops_revenue_2019 %>%
    mutate(revenue = `Amount (Per 1000)`, .by = State) %>%   
    ggplot(aes(x = long, y = lat)) + 
    geom_polygon(aes(group=group, fill = revenue / population)) +
    scale_fill_continuous(name = "Total State Revenue ($1000/Capita)") +
    ggtitle("State Total Revenue (2019)")

highwayFats_states_pops_revenue_2019 %>% ggplot(aes(x=`Amount (Per 1000)`/population, y=`TotalFatalities`/population)) + 
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  ggtitle("State Revenue's Effect on Traffic Fatalities (2019)") +
  xlab("Total State Revenue ($1000/Per Capita)") + ylab("Total Traffic Fatalities Per Capita")
```

#### Investigating Correlations of Traffic Fatalities with Police Expenditure

```{r}
highwayFats_states_pops_revenue_2019 <-
    fat_states_pops_fin_2019 %>% 
    filter(RevenueType == "Police protection", FatalityType == "Highway Fatalities") %>% 
    select(group, long, lat, State, FatalityType, TotalFatalities, `Amount (Per 1000)`, population) %>%
    distinct()

highwayFats_states_pops_revenue_2019 %>% ggplot(aes(x=`Amount (Per 1000)`/population, y=`TotalFatalities`/population)) + 
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  ggtitle("State Police Funding's Effect on Traffic Fatalities (2019)") +
  xlab("Police Funding ($1000/Per Capita)") + ylab("Total Traffic Fatalities Per Capita")
```

#### Investigating Correlations of Traffic Fatalities with Total State Expenditure

```{r}
highwayFats_states_pops_revenue_2019 <-
    fat_states_pops_fin_2019 %>% 
    filter(RevenueType == "Total expenditure", FatalityType == "Highway Fatalities") %>% 
    select(group, long, lat, State, FatalityType, TotalFatalities, `Amount (Per 1000)`, population) %>%
    distinct()

highwayFats_states_pops_revenue_2019 %>%
    mutate(revenue = `Amount (Per 1000)`, .by = State) %>%   
    ggplot(aes(x = long, y = lat)) + 
    geom_polygon(aes(group=group, fill = revenue / population)) +
    scale_fill_continuous(name = "Total State Expenditure ($1000/Capita)") +
    ggtitle("State Expenditure's (2019)")

highwayFats_states_pops_revenue_2019 %>% ggplot(aes(x=`Amount (Per 1000)`/population, y=`TotalFatalities`/population)) + 
  geom_point() +
  geom_smooth(method=lm, se=FALSE)  +
  ggtitle("State Expenditure's Effect on Traffic Fatalities (2019)") +
  xlab("Total Expenditure ($1000/Per Capita)") + ylab("Total Traffic Fatalities Per Capita")
```

#### Investigating Correlations of Traffic Fatalities with Education Expenditure

```{r}
highwayFats_states_pops_revenue_2019 <-
    fat_states_pops_fin_2019 %>% 
    filter(RevenueType == "Education", FatalityType == "Highway Fatalities") %>% 
    select(group, long, lat, State, FatalityType, TotalFatalities, `Amount (Per 1000)`, population) %>%
    distinct()

highwayFats_states_pops_revenue_2019 %>% ggplot(aes(x=`Amount (Per 1000)`/population, y=`TotalFatalities`/population)) + 
  geom_point() +
  geom_smooth(method=lm, se=FALSE)  +
  ggtitle("State Education Funding's Effect on Traffic Fatalities (2019)") +
  xlab("Education Funding ($1000/Per Capita)") + ylab("Total Traffic Fatalities Per Capita")
```

### Results:

#### Analysis of COVID and Fatality Impact 2020-2021

The first cases of COVID-19 reached the US in January 2020, and the
Trump Administration declared a nationwide emergency on the COVID-19
pandemic in March 2021.
(<https://www.cdc.gov/museum/timeline/covid19.html>).

However, according to the National Highway Traffic Safety
Administration, the number of fatal traffic crashes increased by 6.8%.
This would seem strange, considering that less people overall were
driving on the road without the need to commute to work.
(<https://www.nhtsa.gov/press-releases/2020-traffic-crash-data-fatalities>)

Megan McArdle, a columnist the Washington Post, indicated that this may
be due to an increase in young and risky male drivers taking advantage
of empty COVID-era roads. A study done by the AAA Foundation for Traffic
Safety indicated that young males were the only people who indicated
that they were driving more during the COVID pandemic - taking their
risky driving with them.
(<https://www.washingtonpost.com/opinions/2022/03/10/pandemic-risky-driving-maybe-here-to-stay/>).

In addition, according to Michael Wilbur, a PHD Student in Vanderbilt
University, in a paper published in the Transportation Research Record:
Journal of the Transportation Research Board, bus fatalities is also an
exception to the sudden spike of fatal accidents due to a decrease in
operation hours for buses during the COVID-19 pandemic, as well as
social distancing measures and precautions advised and enforced by state
legislature. (<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10107069/>)

#### Deadliest States to Live In's Correlation with Traffic Fatalities Per Capita

Christy Bieber, J.D., indicated that the top 5 most deadliest states to
drive in are Montana, South Carolina, New Mexico, Wyoming, and Texas
based on data from the National Highway Traffic Safety Administration
(NHTSA). This seems to align with our data for Pedestrian Fatalities by
State Population Density.
(<https://www.forbes.com/advisor/legal/most-dangerous-states-for-drivers/>)

#### State Law Enforcement's Impact on Traffic Fatalities

According to Alina Comoreanu, a Senior researcher at WalletHub, the top
5 states that had the most strictest enforcement on speeding and
reckless driving are (in order) Delaware, Colorado, Arizona, New Mexico,
and California. This seems to correlate with our data for Pedestrian
Fatalities by State Population Density.
(<https://wallethub.com/edu/strictest-and-most-lenient-states-on-speeding/14211>)

### Conclusions:

-   We found that an increase in traffic fatalities at the end of the
    reported data (late 2020 and late 2021) during COVID-19 due to
    different factors​

-   Light Trucks and Public Transport were exceptions to this.​

-   We found that the most deadliest states to drive had a correlation
    to high pedestrian fatalities.​

-   We found an inverse correlation between enforcement of speed laws
    and reckless driving and traffic fatalities.​

-   We found Iowa's leading cause of traffic fatalities was highway
    fatalities.​
